---
title: "R-Workshop 2024: RNASeq dataset analysis in R"
author:
  - name: "Dr. Ali T. Abdallah"
    email: "ali.abdallah@uni-koeln.de"
    affiliation: "CECAD Research Center - Bioinformatics Facility"
date: "2024-09-22"
version: "1.0"
site: bookdown::bookdown_site
output:
  rmdformats::readthedown:
    toc_depth: 3
---


This guide walks you through analyzing an RNASeq dataset using R, with a particular focus on the `airway` dataset, which comes prepackaged with R's `airway` library. We'll cover library setup, dataset preprocessing, differential expression analysis, PCA, and visualization techniques such as volcano plots.

---

# 1. Load Libraries and Datasets

Before starting, we need to load various libraries required for RNASeq data analysis and visualization. If these libraries aren't installed yet, R will automatically install them for you.

## Loading and Installing Required Libraries

In R, libraries (also called "packages") are collections of functions and data that extend the basic functionality of R. Before you start analyzing your RNASeq data, you need to install and load several specialized libraries. These libraries provide tools for handling biological data, visualizing results, and performing statistical analysis.

Let's walk through the process of installing and loading the necessary libraries.

### 1. Checking if Libraries Are Installed and Installing Them if Missing

```{r}
# Check if the BiocManager package is installed, if not, install it
if(!requireNamespace("BiocManager", quietly = TRUE)) 
  install.packages("BiocManager")
```

**Explanation:**
- The `requireNamespace()` function checks whether a package is installed. 
- The `quietly = TRUE` argument suppresses messages during the check.
- If the package is not installed, the `install.packages()` function installs it. In this case, we are installing `BiocManager`, a tool that helps you install Bioconductor packages, which are commonly used in bioinformatics.

---

```{r}
# Install DESeq2 if it's not already installed
if(!requireNamespace("DESeq2", quietly = TRUE)) 
  BiocManager::install("DESeq2")
```

**Explanation:**
- `DESeq2` is a powerful package designed for differential gene expression analysis. It’s often used to analyze RNASeq data to find genes that are upregulated or downregulated between conditions.
- If the package is not found on your system, the `BiocManager::install()` function installs it from the Bioconductor repository.

---

```{r}
# Install dplyr if it's not already installed
if(!requireNamespace("dplyr", quietly = TRUE)) 
  install.packages("dplyr")
```

**Explanation:**
- `dplyr` is a widely-used package for data manipulation. You will use it to clean and manipulate your RNASeq data, such as filtering genes or organizing metadata.
- This package is part of the tidyverse, a popular set of tools for data science in R.

---

```{r}
# Install the org.Hs.eg.db package for human gene annotations
if(!requireNamespace("org.Hs.eg.db", quietly = TRUE)) 
  BiocManager::install("org.Hs.eg.db")
```

**Explanation:**
- `org.Hs.eg.db` provides mappings between Ensembl gene IDs and gene symbols for **Homo sapiens (human)**. You will need this package to convert gene identifiers (often used in RNASeq) to human gene symbols.

---

```{r}
# Install airway package if not already installed (RNASeq dataset)
if(!requireNamespace("airway", quietly = TRUE)) 
  BiocManager::install("airway")
```

**Explanation:**
- The `airway` package contains the RNASeq dataset that will be used throughout this tutorial. It includes gene expression data for human airway smooth muscle cells treated with dexamethasone.

---

```{r}
# Install ggplot2 if it's not already installed
if(!requireNamespace("ggplot2", quietly = TRUE)) 
  install.packages("ggplot2")
```

**Explanation:**
- `ggplot2` is a powerful and flexible package for data visualization. It will be used to create plots such as PCA (Principal Component Analysis) plots and volcano plots to visualize your RNASeq analysis results.

---

```{r}
# Install ggrepel for enhanced plotting (avoids label overlap)
if(!requireNamespace("ggrepel", quietly = TRUE)) 
  install.packages("ggrepel")
```

**Explanation:**
- `ggrepel` enhances `ggplot2` plots by preventing text labels from overlapping. It is especially useful in creating clear and readable plots for labeling significant genes in RNASeq analysis.

---

```{r}
# Install DT package for interactive tables
if(!requireNamespace("DT", quietly = TRUE)) 
  install.packages("DT")
```

**Explanation:**
- The `DT` package helps create interactive data tables. You will use it to display the results of your differential expression analysis in a format that allows easy sorting and filtering.

---

### 2. Loading the Installed Libraries

After ensuring that all necessary packages are installed, we need to load them into the current R session. Loading a library makes its functions and data available for use.

```{r message=FALSE, warning=FALSE}
# Load the libraries
library(DESeq2)
library(ggplot2)
library(dplyr)
library(org.Hs.eg.db)
library(airway)
library(ggrepel)
library(DT)
```

**Explanation:**
- This block loads the installed libraries into your R session. Each `library()` call makes the associated package functions and datasets available for immediate use.

---

### 3. Loading the Dataset

Finally, we load the `airway` dataset, which contains RNASeq data from human airway cells.

```{r}
# Load the airway dataset
data(airway)
```

**Explanation:**
- The `data()` function loads built-in datasets from R packages. Here, it loads the `airway` dataset from the `airway` package, which we will use for the RNASeq analysis.

## Explanation
- This code installs the necessary packages if they aren't already available and then loads them for use.
- The `airway` dataset is a sample RNASeq dataset from the `airway` package. It contains gene expression data for studying airway smooth muscle cells under different conditions.

---

# 2. Preprocess the Dataset

Preprocessing involves preparing the RNASeq dataset for analysis, including organizing the feature (gene) data and metadata. You can either load preprocessed files or generate them from scratch.

```{r}
RUN <- F  # Set this to TRUE if you want to preprocess the dataset from scratch
if(RUN){
  my_airway_dataset <- airway
  my_airway_dataset <- assay(my_airway_dataset) %>% as.data.frame()

  # Add gene symbols to the dataset
  my_airway_dataset$gene_symbol <- mapIds(org.Hs.eg.db, keys = rownames(my_airway_dataset), column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")
  
  # Organize feature data (gene symbols and Ensembl gene IDs)
  my_airway_featureData <- my_airway_dataset %>% tibble::add_column(ensembl_gene_id = rownames(.)) %>%
    dplyr::select(ensembl_gene_id, gene_symbol)

  # Clean up the dataset, removing unnecessary columns
  my_airway_dataset <- my_airway_dataset %>% dplyr::select(-gene_symbol)

  # Extract metadata (colData)
  my_airway_colData <- colData(airway) %>% as.data.frame()

  # Save preprocessed data
  saveRDS(my_airway_dataset, file="data/my_airway_dataset.rds")
  saveRDS(my_airway_colData, file="data/my_airway_colData.rds")
  saveRDS(my_airway_featureData, file="data/my_airway_featureData.rds")
} else {
  # Load preprocessed data
  my_airway_dataset <- readRDS("data/my_airway_dataset.rds")
  my_airway_colData <- readRDS("data/my_airway_colData.rds")
  my_airway_featureData <- readRDS("data/my_airway_featureData.rds")
}

```


## Excursus to `SummarizedExperiment`

The class of the airway dataset is:
```{r}
class(airway)
```

In bioinformatics, `SummarizedExperiment` and `RangedSummarizedExperiment` are 
commonly used data structures for storing genomic data. Both handle assay data 
(like expression counts) and metadata, but they differ in their handling of 
genomic coordinates.

---

### SummarizedExperiment

The `SummarizedExperiment` class stores:

- **Assay data** (e.g., gene expression matrix).

- **Row metadata** (e.g., gene names).

- **Column metadata** (e.g., sample information).

However, it does not store genomic ranges (positions on chromosomes).

#### Example:

```{r}
library(SummarizedExperiment)

# Assay data (gene expression counts)
assay_data <- matrix(rnorm(100), nrow=10, ncol=10)
rownames(assay_data) <- paste0("gene", 1:10)
colnames(assay_data) <- paste0("sample", 1:10)

# Metadata
row_data <- data.frame(gene_id=rownames(assay_data))
col_data <- data.frame(condition=c(rep("treated", 5), rep("control", 5)), row.names=colnames(assay_data))

# Create SummarizedExperiment object
se <- SummarizedExperiment(assays = list(counts = assay_data), rowData = row_data, colData = col_data)
se
```

---

### RangedSummarizedExperiment

`RangedSummarizedExperiment` extends `SummarizedExperiment` by including **genomic ranges** (chromosome, start, end, strand) for each gene or region.

#### Example:

```{r}
library(GenomicRanges)
library(SummarizedExperiment)

# Genomic ranges (GRanges object)
gr <- GRanges(seqnames = Rle(c("chr1", "chr1", "chr2", "chr2")),
              ranges = IRanges(start = c(100, 200, 300, 400), end = c(150, 250, 350, 450)),
              strand = Rle(strand(c("+", "-", "+", "-"))))

# Assay data
assay_data <- matrix(rnorm(40), nrow=4, ncol=10)
rownames(assay_data) <- paste0("gene", 1:4)
col_data <- data.frame(condition=c(rep("treated", 5), rep("control", 5)), row.names=colnames(assay_data))

# Create RangedSummarizedExperiment object
rse <- SummarizedExperiment(assays = list(counts = assay_data), rowRanges = gr, colData = col_data)
rse
```

---

### Key Differences

| **Feature**                    | **SummarizedExperiment**                           | **RangedSummarizedExperiment**                     |
|---------------------------------|----------------------------------------------------|----------------------------------------------------|
| **Assay data**                  | Yes                                                | Yes                                                |
| **Row data (gene metadata)**    | Yes, but no genomic coordinates                    | Yes, includes genomic coordinates (via `GRanges`)  |
| **Column data (sample metadata)**| Yes                                                | Yes                                                |
| **Genomic coordinate handling** | No                                                 | Yes (chromosome, start, end, strand)               |

---

### Subsetting by Genomic Ranges

With `RangedSummarizedExperiment`, you can subset data based on genomic positions. For example, subsetting genes on `chr1`:

```{r}
# Subset for genes on chr1
chr1_subset <- rse[rowRanges(rse)$seqnames == "chr1", ]
chr1_subset
```

---

Let us have a look into the data structure of the airway dataset:
```{r}
str(airway)
```
### rowData-function
The rowData-function returns the feature data of the airway dataset. Its class
is:
```{r}
class(rowData(airway))
```

DFrame is a class that is used to store the feature data of the airway dataset.
It is a subclass of DataFrame. The DataFrame class is a subclass of the 
data.frame class. The DataFrame class is used to store data in a tabular format.
The additional functionality provided by DFrame compared to data.frame is
the ability to store metadata. The metadata is stored in the metadata slot of
the DFrame object. The metadata is a list that can store any type of data.
The metadata can be used to store additional information about the data stored
in the DFrame object.

Let us have a look:
```{r}
rowData(airway) %>% head()
```

By using the `str` function you could see the inner world of this DFrame object.
```{r}
str(rowData(airway))
```

As one could see multiple meta-information about the data are present like elementType, elementMetadata, metadata etc. If we compare this to a normal data.frame object
```{r}
str(rowData(airway) %>% as.data.frame())
```
We could clearly see the difference. The DFrame object has additional slots that
could be used to store additional type of information.

### colData-function
The colData-function, called accessor-function, returns the metadata of 
the airway dataset samples. Its class is also a:
```{r}
class(colData(airway))
```

```{r}
str(colData(airway))
```


the content is as follows:
```{r}
colData(airway)
```

### metadata-function

returns a list of objects containing metadata related to overall experiment. 
So it is not directly metadata of the samples as in colData but it adds a 
context to the experiment dataset. The class of the metadata in our case is: 
```{r}
class(metadata(airway)[[1]])
```

MIAME is an acronym for Minimum Information About a Microarray Experiment. 
MIAME is conceptually a set of guidelines for the minimum information that should 
be provided when reporting a microarray experiment. The MIAME guidelines were 
developed by the Microarray Gene Expression Data (MGED) Society. 
The MGED Society is an international organization that aims to promote the sharing 
of microarray data and the development of standards for microarray data.

This framework could be also partially used for RNA-Seq data. The Package Biobase
procide a class MIAME implementing this framework.

```{r}
metadata(airway)[[1]]
```




```{r}
head(my_airway_featureData) %>% as.data.frame()
```



#### Explanation
- **Feature Data**: You extract gene symbols and associate them with Ensembl gene IDs using `org.Hs.eg.db`. This step is critical for adding biological meaning to the analysis.
- **Metadata**: The metadata contains information about the experimental conditions (e.g., `dex` treatment) applied to each sample.
- You can either load preprocessed datasets or preprocess the data on the fly by setting `RUN` to `TRUE`.

---

### Assay-function

Let’s take a quick look at the count data using the assay()-function. It 
expects a SummarizedExperiment object as input and returns the assay data as
a matrix.

```{r}
head(assay(airway))
```

```{r}
class(assay(airway))
```

---


### Why Should Biologists Use `SummarizedExperiment` and `RangedSummarizedExperiment`?

As a biologist diving into bioinformatics and RNASeq data analysis, you may often work with large datasets that contain gene expression levels, sample metadata, and genomic coordinates. Managing these different types of data in a consistent and organized way can be challenging. This is where **`SummarizedExperiment`** and **`RangedSummarizedExperiment`** come in.

#### 1. All Your Data in One Place

Both `SummarizedExperiment` and `RangedSummarizedExperiment` are designed to hold **all the information you need for RNASeq and genomic analysis in one object**:
- **Assay Data**: Your gene expression counts.
- **Row Metadata**: Information about genes or genomic features (like gene IDs or symbols).
- **Column Metadata**: Sample information, such as treatment conditions or time points.

By using these classes, you don't have to manage separate files for each type of data. Everything is stored in a single, well-organized object.

#### 2. Easy to Manipulate and Query

These classes allow you to easily access and manipulate the different types of data:
- You can quickly **filter genes** based on expression levels or metadata.
- You can **subset samples** based on treatment or experimental conditions.
- With `RangedSummarizedExperiment`, you can even **subset by genomic coordinates**, allowing you to focus on specific regions of the genome.

For example, suppose you want to analyze only genes located on a particular chromosome or region. `RangedSummarizedExperiment` makes this kind of genomic subsetting straightforward.

#### 3. Seamless Integration with Bioconductor Tools

Both classes are part of the **Bioconductor** ecosystem, which means they integrate smoothly with many tools designed for biological data analysis. For example:
- The `DESeq2` package for differential expression analysis works directly with these classes.
- Visualization tools like `ggplot2` can easily handle data from these objects.

This means that by learning these classes, you're preparing yourself to work with a wide range of powerful tools in bioinformatics.

#### 4. Reproducibility and Collaboration

Because these objects store data in a consistent format, it's easy to share your analysis with others:
- Your collaborators can load the exact same dataset with all the metadata intact.
- Your workflows become **reproducible**, since everything is stored in one object, minimizing the risk of losing or misplacing parts of the dataset.

#### 5. A Beginner-Friendly Start

Although these classes may sound complex, they are beginner-friendly. Instead of juggling multiple data frames and files, you only need to learn how to work with **one object** that holds all the necessary information. The ability to **access, filter, and modify** the data using simple commands makes it an excellent entry point for biologists starting with R.

---

In summary, `SummarizedExperiment` and `RangedSummarizedExperiment` provide a convenient, organized, and powerful way to manage, analyze, and visualize your biological data in R. They help you focus on the biological questions rather than the data handling, making your research more efficient, reproducible, and scalable.


---

## Filter out lowly expressed genes
```{r}
# Filter out genes with low expression. In this case we require that the sum
# exoression values of each gene across all samples is at least 10.
my_airway_dataset <- airway
my_airway_dataset <- my_airway_dataset[rowSums(assay(my_airway_dataset)) > 10,]
```



# 4. Differential Expression Analysis

The core of RNASeq analysis is identifying genes that are differentially 
expressed under different conditions. In this section, we will use `DESeq2` for 
the analysis.

```{r warning=F, message=F}
# Create a DESeq2 object (DESeqDataSet) from the count matrix and metadata.
dds <- DESeq2::DESeqDataSet(se = my_airway_dataset,
                                      design = ~dex)

# Run the DESeq pipeline for differential expression analysis
dds <- DESeq(dds)
vsd <- vst(dds, blind = T)

# Extract the results table, which contains log2 fold changes and p-values
results <- DESeq2::results(dds) %>% as.data.frame()

# Add significance column based on log2FoldChange and adjusted p-value
results <- results %>% 
  tibble::add_column(significance = 
    if_else(.$log2FoldChange > 1 & .$padj < 0.05, "upregulated", 
    if_else(.$log2FoldChange < -1 & .$padj < 0.05, "downregulated", "not significant")))

# Remove rows with missing values
results <- na.omit(results)

# Add gene symbols to the results using Ensembl gene IDs
results <- results %>% tibble::add_column(ensembl_gene_id = rownames(.)) %>%
  left_join(my_airway_featureData, by = c("ensembl_gene_id"))

# Select relevant columns and arrange by adjusted p-value
results <- results %>% dplyr::select(ensembl_gene_id,
                                     gene_symbol,
                                     baseMean,
                                     log2FoldChange,
                                     padj,
                                     lfcSE,
                                     stat,
                                     pvalue,
                                     significance) %>%
  arrange(padj)

# Display the results in an interactive table
DT::datatable(results, 
              filter = "top",
              extensions = c('Buttons', 'ColReorder'),
              options = list(dom = 'Bfrtip',
                             buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                             colReorder = TRUE))
```

## Explanation
- **DESeq2 Object**: We first create a `DESeqDataSet` object that encapsulates the count data, metadata, and experimental design (in this case, the design is `~dex`, indicating the treatment).
- **Differential Expression**: The `DESeq` function runs the analysis, and we extract results that include fold changes and statistical significance.
- **Interactive Table**: The `DT` package is used to create an interactive table where users can filter and sort results.

---

# 5. Principal Component Analysis (PCA)

PCA is a useful method for visualizing the global structure of high-dimensional datasets like RNASeq.

```{r message=F}
# Transpose the data and convert it into a matrix
data_matrix <- t(assay(vst(dds)))

# Remove columns (features) with zero variance
data_matrix_no_zero_var <- data_matrix[, apply(data_matrix, 2, var) != 0]

# Perform PCA
pca <- prcomp(data_matrix_no_zero_var, scale. = TRUE, center = TRUE, retx = TRUE)

# Extract PCA scores and add metadata
pca_scores <- pca$x %>% as.data.frame() %>% tibble::add_column(dex = colData(dds)$dex)

# Extract PCA loadings (gene contributions)
pca_loadings <- pca$rotation %>% as.data.frame()
pca_loadings <- pca_loadings %>% tibble::add_column(ensembl_gene_id = rownames(.)) %>%
  left_join(my_airway_featureData, by = c("ensembl_gene_id")) %>%
  dplyr::select(ensembl_gene_id, gene_symbol, 1:8)
```

## Explanation
- **Data Preparation**: We transpose the data matrix and remove genes with zero variance to avoid problems in PCA.
- **PCA**: The `prcomp` function performs PCA on the preprocessed data, and the principal components (PCs) are extracted.

---

# 6. Visualization: PCA Plot

Visualize the first two principal components to observe the grouping of samples.

```{r}
# Get PCA summary
sum_pca <- summary(pca)
importance <- sum_pca$importance %>% as.data.frame()

# Create new sample names in the format dex_REP[1:4]
my_airway_colData$Sample.New <- paste0(my_airway_colData$dex,"_REP",rep(1:4, each=2))

pca_scores <- pca_scores %>% tibble::add_column(sample=my_airway_colData$Sample.New)

# Plot the PCA results
ggplot2::ggplot(data=pca_scores, aes(x=PC1, y=PC2, color=dex, label=sample)) + 
  geom_point(size=3) + 
  labs(title="PCA of airway dataset",
       x=paste0("PC1 (", round(100*importance[2,1], 2), "%)"),
       y=paste0("PC2 (", round(100*importance[2,2], 2), "%)"))+
  ggrepel::geom_label_repel()
```

---

# 7. Visualization: Volcano Plot

Volcano plots are a popular method for visualizing fold changes against significance.

## Initial Volcano Plot

```{r}
vp <- ggplot2::ggplot(data = results, aes(x = log2FoldChange, y = -log10(padj), color = significance)) + 
  geom_point() + 
  ggtitle("Volcano plot of the airway dataset") +
  theme(plot.title = element_text(face = "bold", hjust = .5))

vp
```

## Adjust the Plot

Center the plot and add cutoff lines.

```{r}
# Extract the maximal absolute log2 fold change
max_abs_log2FoldChange <- max(abs(results$log2FoldChange), na.rm = TRUE)
# Extract the maximal -log10(padj) (lowest adjusted p-value)
max_neg_log10_padj <- max(-log10(results$padj), na.rm = TRUE)

# Adjust the plot limits
vp.2 <- vp + xlim(c(-max_abs_log2FoldChange, max_abs_log2FoldChange)) + 
  ylim(c(0, max_neg_log10_padj))

# Add cutoff lines for log2FoldChange and padj
vp.3 <- vp.2 + geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_vline(xintercept = 1, linetype = "dashed")

vp.3
```

## Customize Colors and Highlight Genes

Highlight the top 20 genes by `padj`.

```{r}
custom_colors <- c(upregulated = "darkred", downregulated = "blue", `not significant` = "grey")

vp.4 <- vp.3 + scale_color_manual(values = custom_colors)

# Top 20 genes by adjusted p-value
top_20_genes <- results %>% arrange(padj) %>% head(20)

vp.5 <- vp.4 + 
  geom_text_repel(data = top_20_genes, aes(label = gene_symbol), 
                  box.padding = 0.5, point.padding = 0.5, 
                  segment.color = "grey50", segment.size = 0.5, 
                  segment.alpha = 0.5, max.overlaps = Inf, size = 3)

vp.5
```

---

You now have a comprehensive guide to analyzing RNASeq data in R, including preprocessing, differential expression analysis, PCA, and visualizing the results with plots like the PCA and volcano plots.
